Visualizing Contingency Tables

### 2006-09-14 trunk/vcd/paper/Shadings/JCGS.Rnw
Code:   removed border between cells in legend
Paper:
  finished JCGS revision and rejoinder



git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@562 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

myseed <- 1071
nrep <- 5000
recompute <- TRUE
@

@@

\begin{abstract}
Residual-based shadings for enhancing mosaic and association plots
to visualize independence models for contingency tables are extended
in two directions: (a)~perceptually uniform HCL colors are used and (b)~the
result of an associated significance test is coded by the appearance of
color in the visualization. For obtaining (a), a general strategy for
deriving diverging palettes in the perceptually-based HCL space is suggested. As for (b),
cut offs that control the appearance of color are computed in a data-driven
way based on the conditional permutation distribution of maximum-type
test statistics. The shadings are first established for the case of independence
in 2-way tables and then extended to more general independence models
for multi-way tables, including in particular conditional independence
models.\\[0.3cm]

@@

Relationships between categorical variables are typically
analyzed based on the underlying contingency tables that can be explored
for (in)dependence. Two standard methods from the statistical tool box
are log-linear models \citep[see, e.g.,][]{vcd:Agresti:2002} for modeling
(in)dependence structures and mosaic plots \citep{vcd:Hartigan+Kleiner:1981} for 
bringing them out graphically. Both methods can also be combined such that
a certain mosaic plot visualizes a particular log-linear model by 
controlling splitting order and direction and shading of the tiles
in the mosaic display
\citep{vcd:Friendly:1994,vcd:Friendly:1999,vcd:Theus+Lauer:1999,vcd:Hofmann:2001}.
In particular, \cite{vcd:Friendly:1994} suggested a shading strategy
based on the residuals (typically, Pearson or deviance residuals) of the associated
log-linear model that elevates the mosaic plot from a display for frequencies
in a contingency table to a visualization technique that encompasses both, observed
frequencies and residuals. This shading allows for judging the quality of a model
fit and spotting dependence patterns that have not been accounted for by the model.
Other techniques for visualizing dependence in contingency tables such as
association plots \citep{vcd:Cohen:1980} can also be enhanced by using this
residual-based shading.

@@

The estimated expected cell frequencies under $H_0$ are
$\hat n_{ij} = n_{i+}n_{+j}/n_{++}$.
%FIXME% point out homogeneity problem?
The probably best known and most used
measure of discrepancy between observed and expected values are the Pearson residuals
\begin{equation} \label{eq:Pearson}
r_{ij} \quad = \quad \frac{n_{ij} - \hat n_{ij}}{\sqrt{\hat n_{ij}}}.
\end{equation}
The most convenient way to aggregate the $I \times J$ residuals
to one test statistic is their sum of squares
\begin{equation} \label{eq:ChiSquare}
X^2 \quad = \quad \sum_{i, j}r_{ij}^2,
\end{equation}
because this is known to have an unconditional limiting $\chi^2$ distribution with $(I-1)(J-1)$
degrees of freedom under the null hypothesis.
This is the well-known $\chi^2$ test which is typically introduced first in statistics
textbooks when addressing the independence problem in 2-way tables
\citep[see, e.g.,][]{vcd:Agresti:2002}.

@@

For the arthritis data from Table~\ref{tab:art}, both tests indicate a clearly significant
dependence of improvement on treatment: the sum-of-squares statistic from Equation~\ref{eq:ChiSquare}
is $X^2 = \Sexpr{stat(art_chisq)}$ with a $p$ value of $p = \Sexpr{pval(art_chisq)}$,
and the maximum statistic from Equation~\ref{eq:MaxAbs} is $M = \Sexpr{stat(art_max)}$
with $p = \Sexpr{pval(art_max)}$. Both $p$ values have been computed
from a sample of size $\Sexpr{nrep}$ from the permutation distribution under independence
generated via sampling tables with the same row and column sums $n_{i+}$ and $n_{+j}$
using the \cite{vcd:Patefield:1981} algorithm and computing the maximum statistic for 
each of these tables.

@@

\setkeys{Gin}{width=\textwidth}
\begin{figure}[b!]
\begin{center}
<<classic_plots,fig=TRUE,echo=FALSE,height=6,width=11,results=hide>>=
grid.newpage()
pushViewport(viewport(layout = grid.layout(1, 2)))
pushViewport(viewport(layout.pos.col=1, layout.pos.row=1))
mosaic(art, newpage = FALSE, margins = c(2.5, 4, 2.5, 3))
popViewport()
pushViewport(viewport(layout.pos.col=2, layout.pos.row=1))
assoc(art, newpage = FALSE, margins = c(5, 2, 5, 4))
popViewport(2)
@
\caption{Classic mosaic and association plot for the arthritis data.}
\label{fig:classic}
\end{center}
\end{figure}

@@

\emph{Association plots}
\citep{vcd:Cohen:1980} visualize the table of Pearson residuals:
each cell is represented by a rectangle that has (signed) height
proportional to the corresponding Pearson residual $r_{ij}$ and width proportional to the
square root of the estimated expected counts $\sqrt{\hat n_{ij}}$. Thus, the area is proportional to
the raw residuals $n_{ij} - \hat n_{ij}$. The association plot for the arthritis
data is shown in the right panel of Figure~\ref{fig:classic} which leads to the
same interpretation as the mosaic plot: there are more placebo patients with no 
improvement and fewer with marked improvement than expected under independence---vice
versa for the treated patiens.

@@

The HSV space looks like a cone\footnote{The wikipedia gives a very useful
%FIXME% proper cite for Wikipedia
overview at \url{http://en.wikipedia.org/wiki/HSV_color_space} which also provides
links to comparisons with other color spaces discussed in this paper.}
with black at its peak (zero value) and full color wheels for different saturations at the other end,
around a white center (full value). Type and amount of color are controlled by hue
and saturation, respectively. Typically, polar coordinates $(h, s, v)$ rescaled for $s$ and $v$
to the interval $[0, 100]$ (or the unit interval) are used in this space, giving it the
appearance of a cylinder.
For generating colors in the Friendly shading, the following strategy is used:
The hue $h$ codes the sign of the residuals---$h = 0$ (red hue) is used for negative residuals,
$h = 240$ (blue hue) for positive residuals. The absolute size of the residuals
is then coded by the saturation $s$ which is set to $0$, $50$ and $100$, respectively,
for small/medium/large residuals. The value is always fixed at $v = 100$.
This is also depicted in the upper panel of Figure~\ref{fig:shadingHSVHCL} which shows
the saturation/value plane for the given hues $h = 0$ and $h = 240$. 
The full-color palette shows the
colors used for the Friendly shading when the residuals are increasing from left to right.
The reduced-color palette will be explained in Section~\ref{sec:significance}.

@@

Although this HSV-based shading is already very useful for enhancing mosaic and
association plots and although HSV is a very commonly available implementation of
color spaces, HSV colors in general and the Friendly shading in particular have
a number of disadvantages. Most importantly, HSV colors are not perceptually
uniform because the three HSV dimensions map only poorly to the three
perceptual dimensions of the human visual system  \citep{vcd:Brewer:1999,vcd:Ihaka:2003}.
Consequently, the HSV dimensions are confounded, e.g., saturation
is not uniform across different hues. A fully saturated blue $(240, 100, 100)$ is
perceived to be much darker than a fully saturated red $(0, 100, 100)$ or green $(120, 100, 100)$.
This makes it more difficult for the human eye to judge the size of shaded areas
and can therefore lead to color-caused optical illusions when used in statistical graphs
\citep{vcd:Cleveland+McGill:1983}. Furthermore, flashy fully saturated HSV colors are
good for drawing attention to a plot, but hard to look at for a longer time \citep{vcd:Ihaka:2003}
which makes graphics shaded with such colors harder to interpret. Finally, white is employed as
the neutral color for small residuals in the Friendly shading, however typically
grey is found to convey neutrality or un-interestingness much better than white
\citep{vcd:Brewer:1999}.

@@

In the following, we discuss how ideas similar to those from \cite{vcd:Ihaka:2003}
can be used for deriving diverging HCL palettes that provide a suitable translation
of the ideas from the Friendly shading to perceptually uniform HCL colors. 
The HCL space looks like a distorted double cone with black (zero luminance) at one end and white
at the other (full luminance). In its middle, there is a full color wheel for different
values of chroma (that controls the colorfulness). Unfortunately, the HCL space is not as
regular as the HSV space: although its dimensions are usually also given by a
hue ranging in $[0, 360]$ degrees and chroma and luminance ranging in $[0, 100]$
percent, not all combinations $(h, c, l)$ yield valid HCL colors and the admissable combinations
of $c$ and $l$ vary across different hues $h$. For the task of constructing
a diverging palette, this problem can easily be overcome as we just need two different
hues (a `negative' and a `positive' hue) and hence we can choose two hues that correspond to 
similar shapes in the chroma/luminance plane. The lower panel of Figure~\ref{fig:shadingHSVHCL}
shows two such planes side by side for the hues $h = 0$ and $h = 260$.\footnote{The hue
$h = 260$ is chosen rather than $h = 240$ because its chroma/luminance plane is most similar,
as assessed by the symmetric difference of the planes, to that of $h = 0$.}
To obtain a sequence of colors with the similar properties as the Friendly shading, the palette starts
at a fully saturated red $(0, 100, 50)$, goes via a neutral color, ends at a fully saturated
blue $(260, 100, 50)$, and uses linear interpolation in between. Instead of using white
$(0, 0, 100)$ as the neutral color, a light grey $(0, 0, 90)$ is employed as motivated above.
The diverging palette (see the full-color palette in Figure~\ref{fig:shadingHSVHCL})
uses both chroma---i.e., the colorfulness---and luminance---i.e., 
the amount of grey---to code the absolute size of the quantity visualized---i.e., the
residuals $r_{ij}$---when applied to the independence problem. By changing the neutral
color or by changing the maximum chroma, respectively, this can be changed to using only
chroma or luminance for this purpose, but using both seems to be a very effective way
of visualization and corresponds more closely with the properties of the Friendly shading.

Applying these palettes to the mosaic plot of the arthritis data, yields
the displays in the upper middle and right panels of Figure~\ref{fig:shadings} (with
data-driven cut offs as defined Section~\ref{sec:significance}). This illustrates
that especially the full color cells (in the `marked' column) are less flashy and more
balanced in the HCL shading as compared to the HSV shading.

@@

To see such a sum-of-squares shading in practice, we employ a data on the number of piston
ring failures in three legs (north/center/south) in four different steam-driven compressors
at an FIXME. The contingency table is given in \cite{vcd:Haberman:1973}, re-analyzed
by \cite{vcd:Everitt+Hothorn:2006} and displayed in the lower row of Figure~\ref{fig:shadings}.
Neither the sum-of-squares test ($X^2 = \Sexpr{stat(pring_chisq)}$, $p = \Sexpr{pval(pring_chisq)}$)
nor the maximum test ($M = \Sexpr{stat(pring_max)}$, $p = \Sexpr{pval(pring_max)}$) find
evidence for a departure from independence in the data (at 5\% level). However, as argued
by \cite{vcd:Everitt+Hothorn:2006}, a closer look at the size of the residuals might be
interesting. To do so, we choose two fixed cut offs within the range of residuals, 1 and 1.5,
showing that more failures (than expected under independence) in the center leg and less failures
in the south leg were observed for compressor 1 and vice versa for compressor 4. This is
brought out clearly by all three shadings in the lower row of Figure~\ref{fig:shadings}. 
However, less emphasis is given to this pattern by the sum-of-squares shadings in the
middle and right panel because the reduced color palettes are used due to non-significance
of the associated test (at 5\% level). Comparing the HSV-based and HCL-based version
shows again that the latter are less flashy and more balanced.

@@

How the structure of the independence problem can be exploited for 
selecting a suitable layout and shading for mosaic and association plots,
is exemplified with the conditional independence problem in 3-way tables.
For a table $[n_{ijk}]$ with underlying
theoretical probability distribution $[p_{ijk}]$, this can be formulated as
\begin{equation} \label{eq:H0ci}
H_0: \; \pi_{ij|k} \quad = \quad \pi_{i+|k} \pi_{+j|k}.
\end{equation}
where $\pi_{ij|k}$ are the conditional probabilities given the stratum $k$
with $k = 1, \dots, K$. Under the assumption of conditional independence,
we can again estimate expected frequencies $[\hat n_{ijk}]$ and the corresponding 
residuals $[r_{ijk}]$. To test the conditional independence hypothesis, usually
the sum-of-squares statistic is used
\begin{equation} \label{eq:sumchisq}
\sum_{i, j, k} r_{ijk}^2 \quad = \quad \sum_k X_k^2,
\end{equation}
which is simply the sum of the individual sum-of-squares statistics $X_k^2$ in each stratum $k$.

Alternatively, a maximum statistic similar to that from Equation~\ref{eq:MaxAbs}
can be constructed
\begin{equation} \label{eq:dmax}
\max_{i, j, k} |r_{ijk}| \quad = \quad \max_k M_k.
\end{equation}
As in the 2-way case, this allows for identification of the cells that are responsible
for the deviation from conditional independence (if any). If it is not so much of interest
in which \emph{cell} but only in which \emph{stratum} the deviation occurs, then
it would be natural to use 
\begin{equation} \label{eq:maxchisq}
\max_k \sum_{i,j} r_{ijk}^2 \quad = \quad \max_k X_k^2.
\end{equation}
Given a critical value for this statistic, all strata $k$ whose associated sum-of-squares
statistics $X_k^2$ exceed the critical value are in conflict with the hypothesis of 
conditional independence.

All these statistics are of type
\begin{equation} \label{eq:coindep_test}
\lambda_\mathrm{agg} (\lambda_\mathrm{indep}([r_{ijk}])),
\end{equation}
where $\lambda_\mathrm{indep}$ is a functional for assessing independence in stratum $k$
and $\lambda_\mathrm{agg}$ is a functional for aggregating over the $k = 1, \dots, K$ strata.
If the maximum is used for the latter, then identification of the strata responsible for the
dependence is possible. If additionally $\lambda_\mathrm{indep}$ is the maximum,
the corresponding cells can also be identified. Hence, the double maximum
statistic from Equation~\ref{eq:dmax} is the only functional allowing for detection of
both the strata and the cells violating the conditional independence hypothesis.

@@

For illustration, a 3-way and a 4-way table are employed: The first is taken
from a case-control study of smoking and {A}lzheimer's disease published in
\cite{vcd:Salib+Hillier:1997} and re-analyzed using conditional inference techniques
in \cite{vcd:Hothorn+Hornik+VanDeWiel:2006}. It provides data on the smoking behaviour
(no, $<$10, 10--20, $>$20 cigarettes per day), disease status (Alzheimer's, other
dementias, other diagnoses) and gender. The question is whether smoking and disease 
status are conditionally independent given gender. All three statistics
suggested above find evidence for departure from significance: the sum-of-squares
statistic from Equation~\ref{eq:sumchisq} is $\sum_k X_k^2 = \Sexpr{stat(alz_sumchisq)}$
($p = \Sexpr{pval(alz_sumchisq)}$), the maximum sum-of-squares statistic from
Equation~\ref{eq:maxchisq} is $\max_k X_k^2 = \Sexpr{stat(alz_maxchisq)}$
($p = \Sexpr{pval(alz_maxchisq)}$), and the double maximum statistic from Equation~\ref{eq:dmax}
is $\max_k M_k = \Sexpr{stat(alz_dmax)}$ ($p = \Sexpr{pval(alz_dmax)}$).
The $p$ values are again computed by drawing $\Sexpr{nrep}$ samples from the corresponding
permutation distribution. The conditional mosaic plot in Figure~\ref{fig:alz}
shows clearly that the association of smoking and disease is present only in
the group of male patients. The double maximum shading employed allows for
identification of the male heavy smokers as the cells `responsible' for the dependence:
other dementias are more frequent and Alzheimer's disease less frequent in this group than expected
under independence. Interestingly, there seems to be another large residual for
the light smoker group ($<$10 cigarettes) and Alzheimer's disease---however, this
is only significant at 10\% and not at the 1\% level as the other two cells.

@@

As a 4-way example, we use the punishment data \cite{vcd:andersen:1991} 
from a study of the Gallup Institute in Denmark in 1979 about
the attitude of a random sample of 783 persons towards corporal
punishment of children. The contingency table comprises four margins:
memory of punishments as a child (yes/no), attitude as a binary variable
(approval of ``moderate'' punishment or ``no'' approval), highest level
of education (elementary/secondary/high), and age group (15--24, 25--39, $\ge$40 years).
It is of interest whether there is an association between memories of corporal punishments
as a child and attitude towards punishment of children as an adult, controlling
for age and education. Figure~\ref{fig:pun} shows a conditional association
plot of memory and attitude given age and education. This can be
interpreted as a diagnostic residual plot for the associated log-linear conditional
indpendence model. It shows an association between memories and attitude
for the lowest education group (first column) and highest age group (last row):
experienced violence seems to engender violence again as there are less adults
that disapprove punishment in the group with memories of punishments than expected
under independence. For the remaining four age-education groups, there seems to be
no association: all residuals of the conditional independence model are very close to
zero in these cells. All three tests agree again that there is significant
departure from conditional independence in this table: the sum-of-squares
statistic is $\sum_k X_k^2 = \Sexpr{stat(pun_sumchisq)}$ ($p = \Sexpr{pval(pun_sumchisq)}$),
the maximum sum-of-squares statistic is $\max_k X_k^2 = \Sexpr{stat(pun_maxchisq)}$
($p = \Sexpr{pval(pun_maxchisq)}$), and the double maximum statistic
is $\max_k M_k = \Sexpr{stat(pun_dmax)}$ ($p = \Sexpr{pval(pun_dmax)}$). The $\max_k X_k^2$
result is visualized by means of a maximum sum-of-squares shading in Figure~\ref{fig:pun}
with user-defined cut offs 1 and 2, chosen to be within the range of the residuals.
The full-color palette is used only for those strata associated with a
sum-of-squares statistic $X_k^2$ significant at (overall) 5\% level, the reduced-color
palette is used otherwise. This highlights that the dependence pattern is significant
only for the middle and high age group in the low education column. The other panels
in the first column and last row also show a similar dependence pattern, however, it is not
significant at 5\% level and hence graphically down-weighted by using reduced color.

>>>>

myseed <- 1071
nrep <- 5000
nrep_char <- "5,000"
recompute <- TRUE
@

@@

\begin{abstract}
Residual-based shadings for enhancing mosaic and association plots
to visualize independence models for contingency tables are extended
in two directions: (a)~perceptually uniform Hue-Chroma-Luminance colors are used and (b)~the
result of an associated significance test is coded by the appearance of
color in the visualization. For obtaining (a), a general strategy for
deriving diverging palettes in the perceptually-based HCL space is suggested. As for (b),
cut offs that control the appearance of color are computed in a data-driven
way based on the conditional permutation distribution of maximum-type
test statistics. The shadings are first established for the case of independence
in 2-way tables and then extended to more general independence models
for multi-way tables, including in particular conditional independence
models.\\[0.3cm]

@@

Relationships between categorical variables are typically
analyzed based on the underlying contingency tables that can be explored
for (in)dependence. Two standard methods from the statistical tool box
are log-linear models \citep[see, e.g.,][]{vcd:Agresti:2002} for modeling
(in)dependence structures and mosaic plots \citep{vcd:Hartigan+Kleiner:1981} for 
bringing them out graphically. Both methods can also be combined such that
a certain mosaic plot visualizes a particular log-linear model by 
controlling splitting order and direction and shading of the tiles
in the mosaic display
\citep{vcd:Friendly:1994,vcd:Friendly:1999,vcd:Theus+Lauer:1999,vcd:Hofmann:2001}.
In particular, \cite{vcd:Friendly:1994} suggested a shading strategy
based on the residuals (typically, Pearson or deviance residuals) of the associated
log-linear model that elevates the mosaic plot from a display for frequencies
in a contingency table to a visualization technique that encompasses both observed
frequencies and residuals. This shading allows for judging the quality of a model
fit and spotting dependence patterns that have not been accounted for by the model.
Other techniques for visualizing dependence in contingency tables such as
association plots \citep{vcd:Cohen:1980} can also be enhanced by using this
residual-based shading.

@@

The estimated expected cell frequencies under $H_0$ are
$\hat n_{ij} = n_{i+}n_{+j}/n_{++}$. As well-established in the stastical literature,
a very closely related hypothesis is that of homogeneity which in particular leads to the same
expected cell frequencies and is hence not discussed explicitely below.
The probably best known and most used
measure of discrepancy between observed and expected values are the Pearson residuals
\begin{equation} \label{eq:Pearson}
r_{ij} \quad = \quad \frac{n_{ij} - \hat n_{ij}}{\sqrt{\hat n_{ij}}}.
\end{equation}
The most convenient way to aggregate the $I \times J$ residuals
to one test statistic is their sum of squares
\begin{equation} \label{eq:ChiSquare}
X^2 \quad = \quad \sum_{i, j}r_{ij}^2,
\end{equation}
because this is known to have an unconditional limiting $\chi^2$ distribution with $(I-1)(J-1)$
degrees of freedom under the null hypothesis.
This is the well-known $\chi^2$ test which is typically introduced first in statistics
textbooks when addressing the independence problem in 2-way tables
\citep[see, e.g.,][]{vcd:Agresti:2002}.

@@

For the arthritis data from Table~\ref{tab:art}, both tests indicate a clearly significant
dependence of improvement on treatment: the sum-of-squares statistic from Equation~\ref{eq:ChiSquare}
is $X^2 = \Sexpr{stat(art_chisq)}$ with a $p$ value of $p = \Sexpr{pval(art_chisq)}$,
and the maximum statistic from Equation~\ref{eq:MaxAbs} is $M = \Sexpr{stat(art_max)}$
with $p = \Sexpr{pval(art_max)}$. Both $p$ values have been computed
from a sample of size $\Sexpr{nrep_char}$ from the permutation distribution under independence
generated via sampling tables with the same row and column sums $n_{i+}$ and $n_{+j}$
using the \cite{vcd:Patefield:1981} algorithm and computing the respective statistic for 
each of these tables.

@@

@@

\emph{Association plots}
\citep{vcd:Cohen:1980} visualize the table of Pearson residuals:
each cell is represented by a rectangle that has (signed) height
proportional to the corresponding Pearson residual $r_{ij}$ and width proportional to the
square root of the estimated expected counts $\sqrt{\hat n_{ij}}$. Thus, the area is proportional to
the raw residuals $n_{ij} - \hat n_{ij}$. The association plot for the arthritis
data is shown in the right panel of Figure~\ref{fig:classic} which leads to the
same interpretation as the mosaic plot: there are more placebo patients with no 
improvement and fewer with marked improvement than expected under independence---vice
versa for the treated patients.

\setkeys{Gin}{width=\textwidth}
\begin{figure}[t!]
\begin{center}
<<classic_plots,fig=TRUE,echo=FALSE,height=6,width=11,results=hide>>=
grid.newpage()
pushViewport(viewport(layout = grid.layout(1, 2)))
pushViewport(viewport(layout.pos.col=1, layout.pos.row=1))
mosaic(art, newpage = FALSE, margins = c(2.5, 4, 2.5, 3))
popViewport()
pushViewport(viewport(layout.pos.col=2, layout.pos.row=1))
assoc(art, newpage = FALSE, margins = c(5, 2, 5, 4))
popViewport(2)
@
\caption{Classic mosaic and association plot for the arthritis data.}
\label{fig:classic}
\end{center}
\end{figure}

@@

The HSV space looks like a cone \citep[see e.g.,][which also provides
links to comparisons with other color spaces discussed in this paper]{vcd:Wiki+HSV:2006}
with black at its peak (zero value) and full color wheels for different saturations at the other end,
around a white center (full value). Type and amount of color are controlled by hue
and saturation, respectively. Typically, polar coordinates $(h, s, v)$ rescaled for $s$ and $v$
to the interval $[0, 100]$ (or the unit interval) are used in this space, giving it the
appearance of a cylinder.
For generating colors in the Friendly shading, the following strategy is used:
The hue $h$ codes the sign of the residuals---$h = 0$ (red hue) is used for negative residuals,
$h = 240$ (blue hue) for positive residuals. The absolute size of the residuals
is then coded by the saturation $s$ which is set to $0$, $50$ and $100$, respectively,
for small/medium/large residuals. The value is always fixed at $v = 100$.
This is also depicted in the upper panel of Figure~\ref{fig:shadingHSVHCL} which shows
the saturation/value plane for the given hues $h = 0$ and $h = 240$. 
The full-color palette shows the
colors used for the Friendly shading when the residuals are increasing from left to right.
The reduced-color palette will be explained in Section~\ref{sec:significance}.

@@

Although this HSV-based shading is already very useful for enhancing mosaic and
association plots and although HSV is a very commonly available implementation of
color spaces, HSV color space in general and the Friendly shading in particular have
a number of disadvantages. Most importantly, HSV colors are not perceptually
uniform because the three HSV dimensions map only poorly to the three
perceptual dimensions of the human visual system  \citep{vcd:Brewer:1999,vcd:Ihaka:2003}.
Consequently, the HSV dimensions are confounded, e.g., saturation
is not uniform across different hues. A fully saturated blue $(240, 100, 100)$ is
perceived to be much darker than a fully saturated red $(0, 100, 100)$ or green $(120, 100, 100)$.
This makes it more difficult for the human eye to judge the size of shaded areas
and can therefore lead to color-caused optical illusions when used in statistical graphs
\citep{vcd:Cleveland+McGill:1983}. Furthermore, flashy fully saturated HSV colors are
good for drawing attention to a plot, but hard to look at for a longer time \citep{vcd:Ihaka:2003}
which makes graphics shaded with such colors harder to interpret. Finally, white is employed as
the neutral color for small residuals in the Friendly shading, however typically
grey is found to convey neutrality or un-interestingness much better than white
\citep{vcd:Brewer:1999}.

@@

In the following, we discuss how ideas similar to those from \cite{vcd:Ihaka:2003}
can be used for deriving diverging HCL palettes that provide a suitable translation
of the ideas from the Friendly shading to perceptually uniform HCL colors. 
The HCL space looks like a distorted double cone with black (zero luminance) at one end and white
at the other (full luminance). In its middle, there is a full color wheel for different
values of chroma (that controls the colorfulness). Unfortunately, the HCL space is not as
regular as the HSV space: although its dimensions are usually also given by a
hue ranging in $[0, 360]$ degrees and chroma and luminance ranging in $[0, 100]$
percent, not all combinations $(h, c, l)$ yield valid HCL colors and the admissable combinations
of $c$ and $l$ vary across different hues $h$. For the task of constructing
a diverging palette, this problem can easily be overcome as we just need two different
hues (a `negative' and a `positive' hue) and hence we can choose two hues that correspond to 
similar shapes in the chroma/luminance plane. The lower panel of Figure~\ref{fig:shadingHSVHCL}
shows two such planes side by side for the hues $h = 0$ and $h = 260$.\footnote{The hue
$h = 260$ is chosen rather than $h = 240$ because its chroma/luminance plane is most similar,
as assessed by the area of the symmetric difference of the planes, to that of $h = 0$.}
To obtain a sequence of colors with the similar properties as the Friendly shading, the palette starts
at a fully saturated red $(0, 100, 50)$, goes via a neutral color, ends at a fully saturated
blue $(260, 100, 50)$, and uses linear interpolation in between. Instead of using white
$(0, 0, 100)$ as the neutral color, a light grey $(0, 0, 90)$ is employed as motivated above.
The diverging palette (see the full-color palette in Figure~\ref{fig:shadingHSVHCL})
uses both chroma---i.e., the colorfulness---and luminance---i.e., 
the amount of grey---to code the absolute size of the quantity visualized---i.e., the
residuals $r_{ij}$---when applied to the independence problem. By changing the neutral
color or by changing the maximum chroma, respectively, this can be changed to using only
chroma or luminance for this purpose, but using both is a very effective way
of visualization (i.e., yields palettes with more distinctive colors)
and corresponds more closely with the properties of the Friendly shading.

Applying these palettes to the mosaic plot of the arthritis data yields
the displays in the upper middle and right panels of Figure~\ref{fig:shadings} (with
data-driven cut offs as defined Section~\ref{sec:significance}). This illustrates
that especially the full color cells (in the `marked' column) are less flashy and more
balanced in the HCL shading as compared to the HSV shading.

@@

To see such a sum-of-squares shading in practice, we employ a data set on the number of piston
ring failures in three legs (north/center/south) in four different steam-driven compressors
at an Imperial Chemical Industries plant. The contingency table is given in \cite{vcd:Haberman:1973}, re-analyzed
by \cite{vcd:Everitt+Hothorn:2006} and displayed in the lower row of Figure~\ref{fig:shadings}.
Neither the sum-of-squares test ($X^2 = \Sexpr{stat(pring_chisq)}$, $p = \Sexpr{pval(pring_chisq)}$)
nor the maximum test ($M = \Sexpr{stat(pring_max)}$, $p = \Sexpr{pval(pring_max)}$) find
evidence for a departure from independence in the data (at the 5\% level). However, as argued
by \cite{vcd:Everitt+Hothorn:2006}, a closer look at the size of the residuals might be
interesting. To do so, we choose two fixed cut offs within the range of residuals, 1 and 1.5,
showing that more failures (than expected under independence) in the center leg and less failures
in the south leg were observed for compressor 1 and vice versa for compressor 4. This is
brought out clearly by all three shadings in the lower row of Figure~\ref{fig:shadings}. 
However, less emphasis is given to this pattern by the sum-of-squares shadings in the
middle and right panel because the reduced color palettes are used due to non-significance
of the associated test (at 5\% level). Comparing the HSV-based and HCL-based version
shows that the latter are less flashy and more balanced.

@@

How the structure of the independence problem can be exploited for 
selecting a suitable layout and shading for mosaic and association plots,
is exemplified with the conditional independence problem in 3-way tables.
For a table $[n_{ijk}]$ with underlying
theoretical probability distribution $[\pi_{ijk}]$, this can be formulated as
\begin{equation} \label{eq:H0ci}
H_0: \; \pi_{ij|k} \quad = \quad \pi_{i+|k} \pi_{+j|k}.
\end{equation}
where $\pi_{ij|k}$ are the conditional probabilities given the stratum $k$
with $k = 1, \dots, K$. Under the assumption of conditional independence,
we can again estimate expected frequencies $[\hat n_{ijk}]$ and the corresponding 
residuals $[r_{ijk}]$. To test the conditional independence hypothesis, usually
the sum-of-squares statistic is used
\begin{equation} \label{eq:sumchisq}
\sum_{i, j, k} r_{ijk}^2 \quad = \quad \sum_k X_k^2,
\end{equation}
which is simply the sum of the individual sum-of-squares statistics $X_k^2$ in each stratum $k$.

Alternatively, a maximum statistic similar to that from Equation~\ref{eq:MaxAbs}
can be constructed
\begin{equation} \label{eq:dmax}
\max_{i, j, k} |r_{ijk}| \quad = \quad \max_k M_k.
\end{equation}
As in the 2-way case, this allows for identification of the cells which are responsible
for the deviation from conditional independence (if any). If it is not so much of interest
in which \emph{cell} but only in which \emph{stratum} the deviation occurs, then
it would be natural to use 
\begin{equation} \label{eq:maxchisq}
\max_k \sum_{i,j} r_{ijk}^2 \quad = \quad \max_k X_k^2.
\end{equation}
Given a critical value for this statistic, all strata $k$ whose associated sum-of-squares
statistics $X_k^2$ exceed the critical value are in conflict with the hypothesis of 
conditional independence.

All these statistics are of type
\begin{equation} \label{eq:coindep_test}
\lambda_\mathrm{agg} (\lambda_\mathrm{indep}([r_{ijk}])),
\end{equation}
where $\lambda_\mathrm{indep}$ is a functional for assessing independence in stratum $k$
and $\lambda_\mathrm{agg}$ is a functional for aggregating over the $k = 1, \dots, K$ strata.
If the maximum is used for the latter, then identification of the strata responsible for the
non-independence is possible. If additionally $\lambda_\mathrm{indep}$ is the maximum,
the corresponding cells can also be identified. Hence, the double maximum
statistic from Equation~\ref{eq:dmax} is the only functional allowing for detection of
both the strata and the cells violating the conditional independence hypothesis.

@@

For illustration, a 3-way and a 4-way table are employed: The first is taken
from a case-control study of smoking and {A}lzheimer's disease published in
\cite{vcd:Salib+Hillier:1997} and re-analyzed using conditional inference techniques
in \cite{vcd:Hothorn+Hornik+VanDeWiel:2006}. It provides data on the smoking behaviour
(no, $<$10, 10--20, $>$20 cigarettes per day), disease status (Alzheimer's, other
dementias, other diagnoses) and gender. The question is whether smoking and disease 
status are conditionally independent given gender. All three statistics
suggested above find evidence for departure from significance: the sum-of-squares
statistic from Equation~\ref{eq:sumchisq} is $\sum_k X_k^2 = \Sexpr{stat(alz_sumchisq)}$
($p = \Sexpr{pval(alz_sumchisq)}$), the maximum sum-of-squares statistic from
Equation~\ref{eq:maxchisq} is $\max_k X_k^2 = \Sexpr{stat(alz_maxchisq)}$
($p = \Sexpr{pval(alz_maxchisq)}$), and the double maximum statistic from Equation~\ref{eq:dmax}
is $\max_k M_k = \Sexpr{stat(alz_dmax)}$ ($p = \Sexpr{pval(alz_dmax)}$).
The $p$ values are again computed by drawing $\Sexpr{nrep_char}$ samples from the corresponding
permutation distribution. The conditional mosaic plot in Figure~\ref{fig:alz}
shows clearly that the association of smoking and disease is present only in
the group of male patients. The double maximum shading employed allows for
identification of the male heavy smokers as the cells `responsible' for the dependence:
other dementias are more frequent and Alzheimer's disease less frequent in this group than expected
under independence. Interestingly, there seems to be another large residual for
the light smoker group ($<$10 cigarettes) and Alzheimer's disease---however, this
is only significant at 10\% and not at the 1\% level as the other two cells.

@@

As a 4-way example, we use the punishment data \citep{vcd:andersen:1991} 
from a study of the Gallup Institute in Denmark in 1979 about
the attitude of a random sample of 783 persons towards corporal
punishment of children. The contingency table comprises four margins:
memory of punishments as a child (yes/no), attitude as a binary variable
(approval of ``moderate'' punishment or ``no'' approval), highest level
of education (elementary/secondary/high), and age group (15--24, 25--39, $\ge$40 years).
It is of interest whether there is an association between memories of corporal punishments
as a child and attitude towards punishment of children as an adult, controlling
for age and education. Figure~\ref{fig:pun} shows a conditional association
plot of memory and attitude given age and education. This can be
interpreted as a diagnostic residual plot for the associated log-linear conditional
independence model. It shows an association between memories and attitude
for the lowest education group (first column) and highest age group (last row):
experienced violence seems to engender violence again as there are less adults
that disapprove punishment in the group with memories of punishments than expected
under independence. For the remaining four age-education groups, there seems to be
no association: all residuals of the conditional independence model are very close to
zero in these cells. All three tests agree again that there is significant
departure from conditional independence in this table: the sum-of-squares
statistic is $\sum_k X_k^2 = \Sexpr{stat(pun_sumchisq)}$ ($p = \Sexpr{pval(pun_sumchisq)}$),
the maximum sum-of-squares statistic is $\max_k X_k^2 = \Sexpr{stat(pun_maxchisq)}$
($p = \Sexpr{pval(pun_maxchisq)}$), and the double maximum statistic
is $\max_k M_k = \Sexpr{stat(pun_dmax)}$ ($p = \Sexpr{pval(pun_dmax)}$). The $\max_k X_k^2$
result is visualized by means of a maximum sum-of-squares shading in Figure~\ref{fig:pun}
with user-defined cut offs 1 and 2, chosen to be within the range of the residuals.
The full-color palette is used only for those strata associated with a
sum-of-squares statistic $X_k^2$ significant at (overall) 5\% level, the reduced-color
palette is used otherwise. This highlights that the dependence pattern is significant
only for the middle and high age group in the low education column. The other panels
in the first column and last row also show a similar dependence pattern, however, it is not
significant at 5\% level and hence graphically down-weighted by using reduced color.
